---
title: "Homework 3"
subtitle: "Due Wednesday September 18, 2019"
author: "Eric Bae"
date: '`r Sys.Date()`'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(knitr)
library(tidyr)
library(dplyr)
```

For each assignment, turn in by the due date/time.  Late assignments must be arranged prior to submission.  In every case, assignments are to be typed neatly using proper English in Markdown.  

This week, we spoke about R and version control, munging and 'tidying' data, good programming practice and finally some basic programming building blocs.  To begin the homework, we will for the rest of the course, start by loading data and then creating tidy data sets.  

## Problem 1

Work through the "Getting and Cleaning Data" lesson parts 3 and 4.

From the R command prompt:  
```{r eval=FALSE}
library(swirl)  
swirl()  
```

## Problem 2

Create a new R Markdown file within your local GitHub repo folder (file-->new-->R Markdown-->save as).

The filename should be: HW3_lastname, i.e. for me it would be HW3_Settlage

You will use this new R Markdown file to solve the following problems.

  
## Problem 3

Redo Problem 4 parts a-d from last time using the tidyverse functions and piping.

a. Sensory data from five operators.    
<http://www2.isye.gatech.edu/~jeffwu/wuhamadabook/data/Sensory.dat> 
b. Gold Medal performance for Olympic Men's Long Jump, year is coded as 1900=0.  
<http://www2.isye.gatech.edu/~jeffwu/wuhamadabook/data/LongJumpData.dat>  
c. Brain weight (g) and body weight (kg) for 62 species.    
<http://www2.isye.gatech.edu/~jeffwu/wuhamadabook/data/BrainandBodyWeight.dat>  
d. Triplicate measurements of tomato yield for two varieties of tomatos at three planting densities.  
<http://www2.isye.gatech.edu/~jeffwu/wuhamadabook/data/tomato.dat>  

&nbsp; 

```{r}
# Sensory Table
sensory.url <- "http://www2.isye.gatech.edu/~jeffwu/wuhamadabook/data/Sensory.dat"

# Importing data as dataframe
sensory <- as.data.frame(fread(sensory.url, fill = TRUE, skip = 2))
N <- nrow(sensory) # Number of rows = 30
D <- ncol(sensory) - 1 # Number of objectives/variables = 5

# Rearranging data so that NA goes to the first row
for (i in 1:N) {
  if (is.na(sensory$V6[i])) {
    sensory[i,-1] <- sensory[i,1:D]
    sensory[i,1] <- NA
  }
}

# Total number of items = 10
I <- length(sensory$V1[which(is.na(sensory$V1) == FALSE)])

# Reorganizes item
sensory$V1 <- rep(1:I, each = 3)

# Tidying up, one column by column
Item <- sort(rep(sensory$V1, each = D))
#Observation <- rep(rep(1:3, each = 5), I)
Operator <- rep(rep(1:D), 3*I)
Dat <- c(t(sensory[,-1]))

# Combining the columns
#sensory <- as_tibble(cbind(Item, Observation, Operator, Dat))
sensory <- as_tibble(cbind(Item, Operator, Dat))

# Summary statistics
summary(sensory)
sensory.operator <- sensory %>%
  group_by(Operator) %>%
  summarize(Mean = mean(Dat), SD = sd(Dat), Min = min(Dat), Max = max(Dat))
sensory.item <- sensory %>%
  group_by(Item) %>%
  summarize(Mean = mean(Dat), SD = sd(Dat), Min = min(Dat), Max = max(Dat))
sensory.itemoperator <- sensory %>%
  group_by(Item, Operator) %>%
  summarize(Mean = mean(Dat), SD = sd(Dat), Min = min(Dat), Max = max(Dat))
sensory.operator
sensory.item
sensory.itemoperator
```

\fbox{\begin{minipage}{\linewidth}
The summary statistics shows that the all-time minimum sensory measurement (no idea what the unit is) is 0.700, the maximum is 9.400, and the mean is 4.657. 

Grouping by operator, operator 3 had the minimum average measurement with the value of 4.17, and operator 4 had the maximum average measurement with the value of 5.19. 

Grouping by item, item 7 had the minimum mean measurement at 1.41 while item 9 had the highest at 8.47. 
\end{minipage}
}

&nbsp; 

```{r}
# Kable
kable(sensory.operator, format = "markdown", 
      caption = "Summary statistics, grouped by operator")
kable(sensory.item, format = "markdown", 
      caption = "Summary statistics, grouped by item")
kable(sensory.itemoperator, format = "markdown", 
      caption = "Summary statistics, grouped by item & operator")

kable(sensory, format = "markdown",  
      caption = "Sensory data, reformated")
```

\fbox{\begin{minipage}{\linewidth}
The above is the tidy version of the data set. There is obviously a lot of issues with this data set. Starting with the fact that it is extremely long, it is hard to navigate and confusing to understand. 

I also had to use a lot of hard-coding, which is not ideal. 
\end{minipage}
}

&nbsp; 

```{r, warning=FALSE}
# Gold Medals Table
medals.url <- "http://www2.isye.gatech.edu/~jeffwu/wuhamadabook/data/LongJumpData.dat"
medals <- as.data.frame(fread(medals.url, fill = TRUE, skip = 1))

# Assign names
colnames(medals) <- rep(c("Year", "Long Jump"), 4)

# Tibble
medals <- as_tibble(rbind(medals[,1:2], medals[,3:4], medals[,5:6], medals[,7:8]))

# Add rankings to my table
medals <- medals %>%
  mutate(Rank = round(rank(-`Long Jump`), 0)) # Added rankings
medals <- medals[-which(is.na(medals$Year)),] # Remove the two NAs

# Summary statistics
medals.summary <- medals %>%
  summarize(Mean = mean(`Long Jump`), SD = sd(`Long Jump`), 
            Min = min(`Long Jump`), Max = max(`Long Jump`))
  
# Kable
kable(medals.summary, format = "markdown", 
      caption = "Summary statistics of Gold Medals data")
kable(medals, format = "markdown", caption = "Gold Medals data, reformated")
```

\fbox{\begin{minipage}{\linewidth}
The above table is the table of the record long jump of the male gold medalist by olympic year. The index year (0) is 1900. I also added the rankings as the third column. 

The minimum record was 249.75, recorded at the 1896 Olympics and the maximum record was 350.50, recorded at the 1968 Olympics. The mean record across the years was 310.3. 

This one went a little bit better than the Sensory data set, though some hard coding was used, especially for rbind().
\end{minipage}
}

&nbsp; 

```{r, warning=FALSE}
# Brain and Body Weight Table
brain.url <- "https://www2.isye.gatech.edu/~jeffwu/wuhamadabook/data/BrainandBodyWeight.dat"
brain <- as.data.frame(fread(brain.url, fill = TRUE, skip = 1))

# Assign names
colnames(brain) <- rep(c("Body Wt", "Brain Wt"), 3)

# As tibble
brain <- as_tibble(rbind.data.frame(brain[,1:2], brain[,3:4], brain[,5:6]))
brain <- brain[-which(is.na(brain$`Body Wt`)),]

# Add brain weight/body weight ratio
brain <- brain %>%
  arrange(`Body Wt`) %>%
  mutate(`Brain-to-Body Ratio` = `Brain Wt`/`Body Wt`)

# Summary statistics
brain.bodywt <- brain %>%
  summarize(`Body Wt Mean` = mean(`Body Wt`), 
            `Body Wt SD` = sd(`Body Wt`), 
            `Body Wt Min` = min(`Body Wt`), 
            `Body Wt Max` = max(`Body Wt`))
brain.brainwt <- brain %>%
  summarize(`Brain Wt Mean` = mean(`Brain Wt`), 
            `Brain Wt SD` = sd(`Brain Wt`), 
            `Brain Wt Min` = min(`Brain Wt`), 
            `Brain Wt Max` = max(`Brain Wt`))

# Kable
kable(brain.brainwt, format = "markdown", 
      caption = "Summary statistics of brain wt")
kable(brain.bodywt, format = "markdown", 
      caption = "Summary statistics of body wt")
kable(brain, format = "markdown", caption = "Brain and Body Weight data, reformated")
```

\fbox{\begin{minipage}{\linewidth}
The average weight of all bodies observed was 198.790 kg whereas the average weight of all brains was 283.13 g. However, the median weight of bodies and brains were 3.342 kg and 17.25 g, respectively, suggesting that both weight were skewed heavily to the right. The weights ranged from 0.005 kg to 6,654.000 kg for body and 0.10 g to 5,712.00 g for brain. 

I added the brain-to-body weight ratio (in 1000s) as the third column just for reference. Though I did not perform statistical analysis, simple eyeballing it seems to indicate that the larger the body weight the lower the ratio. 

The issue here was more similar to the issue I faced with the medals dataset in that there was some hard coding with binding the three pairs of columns into one. 
\end{minipage}
}

&nbsp; 

```{r, warning=FALSE}
# Tomato Yield Table
tomato.url <- "http://www2.isye.gatech.edu/~jeffwu/wuhamadabook/data/tomato.dat"
tomato <- as.data.frame(fread(tomato.url, fill = TRUE, skip = 1))

# Assign names
colnames(tomato) <- as.character(tomato[1, c(4, 1:3)])
colnames(tomato)[1] <- "Tomato Variety"
tomato <- tomato[-1,]
tomato <- tomato[rep(1:2, each = 3),]

new.tomato <- tomato
for (i in seq(1, nrow(tomato), by = 3)) {
  for (j in 2:ncol(tomato)) {
    new.tomato[seq(i, i+2),j] <- strsplit(as.character(tomato[i,j]), ",")[[1]]
  }
}

# Tidying up, one column by column
Tomato <- c(t(rep(new.tomato$`Tomato Variety`, each = ncol(new.tomato) - 1)))
Density <- rep(colnames(new.tomato)[-1], length(unique(Tomato)))
Yield <- as.numeric(c(t(new.tomato[,-1])))
tomato <- as_tibble(cbind.data.frame(Tomato, Density, Yield))

# Summary statistics
tomato.summary <- tomato %>%
  group_by(Tomato) %>%
  summarize(`Mean Tomato Yield` = mean(Yield), 
            `Tomato Yield SD` = sd(Yield), 
            `Min Tomato Yield` = min(Yield), 
            `Max Tomato Yield` = max(Yield))

tomato.summary.den <- tomato %>%
  group_by(Density) %>%
  summarize(`Mean Tomato Yield` = mean(Yield), 
            `Tomato Yield SD` = sd(Yield), 
            `Min Tomato Yield` = min(Yield), 
            `Max Tomato Yield` = max(Yield))

tomato.summary.both <- tomato %>%
  group_by(Tomato, Density) %>%
  summarize(`Mean Tomato Yield` = mean(Yield), 
            `Tomato Yield SD` = sd(Yield), 
            `Min Tomato Yield` = min(Yield), 
            `Max Tomato Yield` = max(Yield))

# Kable
kable(tomato.summary, format = "markdown", 
      caption = "Summary statistics of tomato yield by variety")
kable(tomato.summary.den, format = "markdown", 
      caption = "Summary statistics of tomato yield by density")
kable(tomato.summary.both, format = "markdown", 
      caption = "Summary statistics of tomato yield by variety and density")

kable(tomato, format = "markdown", caption = "Tomato yield by variety, reformated")
```

\fbox{\begin{minipage}{\linewidth}
The first column represents the tomato variety, the second the planting density, and the third the yield. The lowest yield was found in the tomato variety called "PusaEarlyDwarf" and planting density of 10,000, while the highest yield was found in "Ife\#1" and planting density of 30,000. 

I had a lot of trouble with this data set because the yield data was not numeric but actually some sort of characters, delineated by commas. I had to find a way to split them up by the commas, but unable to change the classes of the yields without causing the numbers to change. This is why summary table is not available here because I could not find a way to generate summary statistics for character outputs. 
\end{minipage}
}

&nbsp; 

## Problem 4

Finish this homework by pushing your changes to your repo.  In general, your workflow for this should be:  

1. In terminal: git pull  -- to make sure you have the most recent local repo  
2. In terminal: do some work  
3. In terminal: git add -- check files you want to commit  
4. In terminal: git commit -- make message INFORMATIVE and USEFUL  
5. In terminal: git push -- this pushes your local changes to the repo  

If you have difficulty with steps 1-5, git is not correctly or completely setup.

**Only submit the .Rmd and .pdf solution files.  Names should be formatted HW3_lastname_firstname.Rmd and HW3_lastname_firstname.pdf**

## Optional preperation for next class:  

TBD -- could be something sent as a class message